PhD Title \\ PhD Thesis

### 2024-07-11 Latex_files/Thesis/chapters/chapter 3/main.tex
updates

\chapter{Effects of localised shearing on crystal growth and nucleation}
As outlined in Chapter 1, the original outline of the PhD was to investigate the possibility of using optical tweezing as a means of initiating and controlling nucleation by generating fluid flow within a small droplet of supersaturated solution. The goal of which would be to allow the understand the influence of shearing on nucleation at a micro level as compared to larger scale results. It has been shown that for macro-scale systems, the likelihood of nucleation increases to a maximum value under increased shearing \cite{Debuysschere2023, Mura2016}. Theoretical research into the matter identified two competing effects that effect a crystal in moving fluid fields; firstly, nucleation is enhanced due to the increased mass transfer of solute material; and secondly, shear flow against the crystal surface leading to a decrease in growth \cite{Mura2016}. These two competing effects are validated by experimental work using glycine solutions, showing that beyond a certain shear rate the nucleation rate is reduced \cite{Debuysschere2023}. In this chapter I outline the optical tweezer equipment used during this PhD, the initial attempts made to induce nucleation via optical rotation, and the impact of a moving beam on the crystal growth of a newly formed nucleus.

@@

@@

Vaterite samples where made by the first preparing equal amounts of 
$CaCl_2$ and $Na_2CO_3$ at a concentration of $0.33M$, at the same time
a vial of $0.33M\ MgSO_4$ was prepared and set aside for later. First
a small vial was filled with $1.5mL$ of $CaCl_2$ followed by $60\mu L$ 
and $90\mu L$ of $MgSO_4$ and $NaCO_3$ respectively, forming a seed solution. 
Next, a larger vial was filled with $5\ mL$, $1.5\ mL$, and $1\ mL$ of
$CaCL_2$, $MgSO_4$, and $NaCO_3$ respectively followed by the seed solution. 
After 10 minutes of slow but continuous mixing a few drops of Agepon was added
to halt the reaction, the solution was filtered and washed 3 times with 
distilled water before being suspended in water.

\begin{figure}[h!]
	\centering
	\begin{subfigure}{0.4\linewidth}
		\includegraphics[width=\linewidth]{vaterite_sample.jpg}
		\subcaption{}
	\end{subfigure}
	\begin{subfigure}{0.55\linewidth}
		\includegraphics[width=\linewidth]{rotating_psd.png}
		\subcaption{}
	\end{subfigure}
	\caption{(a)Sample vaterite sphere suspended in water and trapped by circular polarised trap. (b) Collected power spectrum from rotating vaterite, peaks in the power spectrum appear at integer multiples of the rotational frequency ($f_{rot} \approx 49.8$ Hz)}
\end{figure}

Liquid crystals are an intriguing example of materials with mixed phase
properties, unlike typical solutes such as Glycine, a liquid crystal
can still maintain some degree of order between its individual molecules
while in the liquid state. This is due to the fact that liquid crystals
are constructed of long ordered molecules that demonstrate a long range
ordering. There are three main types of liquid crystal transition methods:
thermotropic crystals will transition to their liquid crystal phase when 
sufficiently heated; Lyotropic materials can undergo this transition due 
to changes in temperature and concentration; and lastly Metallotropic 
materials - which are composed of both organic and inorganic molecules - 
change phase according to the ratio of organic to inorganic molecules present.
Liquid crystal rotors are rather simple in their production, 
4-Heptyl-4-biphenylcarbonitrile (7CB) was purchased from Sigma Aldrich
and a small amount was added to a vial of distilled water. The solution 
was then heated in a water bath to $25^\circ$ in order to transition the
solid crystal into its liquid crystal state. The individual droplets 
are inherently birefringent and would rotate readily while suspended in 
distilled water.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.5\linewidth]{LC_sample.png}
	\caption{Liquid crystal undergoing rotation due to the circularly polarised trap.} 
\end{figure}
\subsection{Rotation of Vaterite micro spheres} 
Optical tweezing has often been used for micro-rheology, by computing the
exact forces being exerted on the trapped sphere, one can determine the
local temperature/viscosity of the medium \cite{Millen2014, RodriguezSevilla2018}.
Typically one would use a birefringent particle (i.e. vaterite, liquid
crystals, etc) and rotating it within the fluid, the maximum rotation
rate being a product of the fluid drag resisting the torque of the 
trapping beam \cite{RodriguezSevilla2018}. Or if you want to measure 
fluid flow you can instead use a micro-rotor to see how fluid flow 
propagates in the medium \cite{Knoener2005}. Likewise, one can use 
a galvanometric mirror to probe the drag force of the fluid, by 
understanding the trap strength (calibrating using a low frequency 
signal) one can measure the drag force experienced by the local fluid \cite{RobertsonAnderson2018}. I

@@

From Fig.\ref{fig:vaterite_shear}~(b) it's clear that the rotational speeds
achieved both in the lab and from experimental literature are well below the 
theoretical maximum shear rate reported by \cite{Debuysschere2023} as you move 
away from the surface of your micro-rotor.

\section{Rotation of Vaterite in supersaturated solutions}

Vaterite samples were synthesised according to \cite{Parkin2009, Bishop2004} 
(see sec.~\ref{sec:vaterite}), and then suspended in a droplet of 
supersaturated solutions of Glycine and water. After locating a 
singular microsphere the vaterite was trapped in a circularly 
polarised light and brought close to the droplet edge, after a 
period of ten minutes if no nucleation event was observed the 
particle was released and another particle was trapped. Due to 
the increased viscosity of the supersaturated solution and the 
proximity to the droplet edge, this was done as previous work with
optical tweezers have only seen nucleation near the solution-air interface
\cite{Gowayed2021, Liao2022, Walton2018}. The observed rotation rate 
was rather low or non-existent, the results are tabulated below:
\begin{table}
	\centering
	\caption{Results from rotating vaterite within supersaturated solution of $H_2O$ and Glycine. Solubility concentration for Glycine at $16^\circ$ was $C^*=0.2016g/g$}
	\begin{tabular}[width=\textwidth]{|c|c|c|c|}
		\hline
		Super Saturation & Particle radius [$\mu$] & $\omega$ [Hz] & Nucleation [$\checkmark/\times$]\\
		\hline
		\multirow{3}*{1.01} & 2.34 & 20.7 & $\times$ \\
		\cline{2-4} & 5.67 & 23.3 & $\times$ \\
		\cline{2-4} & 3.26 & 15.4 & $\times$ \\
		\hline
		\multirow{3}*{1.14} & 1.89 & 1.23 & $\times$ \\
		\cline{2-4} & 3.75 & 3.54 & $\times$ \\
		\cline{2-4} & 4.35 & 4.86 & $\times$ \\
		\hline
		\multirow{3}*{1.29} & 3.47 & 0.00 & $\times$ \\
		\cline{2-4} & 1.59 & 0.00 & $\times$ \\
		\cline{2-4} & 6.24 & 0.00 & $\times$ \\
		\hline
		\multirow{3}*{1.40} & 6.32 & 0.00 & $\times$ \\
		\cline{2-4} & 3.68 & 0.00 & $\times$ \\
		\cline{2-4} & 5.43 & 0.00 & $\times$ \\
		\hline
		\multirow{3}*{1.49} & 4.76 & $0.00$ & $\times$ \\
		\cline{2-4} & 7.27 & $0.00$ & $\times$ \\
		\cline{2-4} & 1.52 & $0.00$ & $\times$ \\
		\hline
	\end{tabular}
\end{table}

The calculations from fig.\ref{fig:vaterite_shear} indicate that even
in pure water the rotational speed achieved by a microsphere is
insufficient to achieve a significant shear rate more than a few microns
from its surface. The closest we could trap a microsphere to the droplet
edge was in the range of $5-10\mu\ m$, at that distance the fluid flow
would be so low that not even using a liquid droplet rotor would achieve
the rotational speeds necessary to localise nucleation. While in theory
a sufficiently focused laser could rotate any microsphere to a fast enough 
to match the speeds expected by \cite{Debuysschere2023} the localised intensity
would be so large that even using $D_2O$ would see a significant increase in 
temperature.

\section{Shearing via a Galvano-mirror}

Using a galvano mirror can A particle's motion can be precisely controlled For a simple circular path one can estimate the sphere's speed by the radius of its path and the frequency of its orbit $U = R\omega$; however for a more complex path, such as an elliptical orbit the curve needs to be parametrised. One can describe the position parameter of an ellipse as such:

@@

where a and b are the different characteristic radii of an ellipse, if we assume that u describes time from some initial point we can say $u=t\omega$ where omega is the frequency of orbit. Subbing this in and then taking the partial derivative of position gives:

@@

In order to compute U we simply take the magnitude of our velocity. For low velocities the fluid flow around the entire sphere can be computed based on the sphere's velocity.

@@

>>>>

\chapter{Effects of localised shearing on crystal growth and nucleation}
As outlined in Chapter 1, the original outline of the PhD was to investigate 
the possibility of using optical tweezing as a means of initiating and 
controlling nucleation by generating fluid flow within a small droplet of 
supersaturated solution. The goal of which would be to understand the 
influence of shearing on nucleation at a micro level as compared to larger 
scale results. It has been shown that for macro-scale systems, the likelihood 
of nucleation increases to a maximum value under increased shearing 
\cite{Debuysschere2023, Mura2016}. Theoretical research into the matter 
identified two competing effects that effect a crystal in moving fluid fields; 
firstly, nucleation is enhanced due to the increased mass transfer of solute 
material; and secondly, shear flow against the crystal surface leading to a 
decrease in growth \cite{Mura2016}. These two competing effects are validated 
by experimental work using glycine solutions, showing that beyond a certain 
shear rate the nucleation rate is reduced \cite{Debuysschere2023}. In this 
chapter I outline the optical tweezer equipment used during this PhD, the 
initial attempts made to induce nucleation via optical rotation, and the 
impact of a moving beam on the crystal growth of a newly formed nucleus.

@@

\subsection{Fourier Optics and 4f correlators}
A 4f correlator is an example of Fourier optics in practice, understanding
that a focused lens takes a Fourier transform of the light profile. 
Consider a laser with a circular Gaussian profile, if you were to place 
a detector there you would pick up the intensity as a function of its 
position within the beam. If however you focused the light into a single 
point (using a +ve focal lens) you are actually seeing a measurement of 
the phase of your laser with position, in which you would see a diffraction
limited spot ($d = \lambda/2nsin(\theta)$), indicating that the laser is 
collimated. In imaging systems, a series of focal lenses can be used to 
filter out unwanted scattering from an image (or in an inverse case 
differentiate between different images), the placement of each lens is shown below.
For our applications a 4f correlator is utilised to ensure that the motion
of the galvano-mirrors does not move the focal point of the laser, allowing 
for a stable trap even while in motion.

@@

Vaterite samples where made by the first preparing equal amounts of 
$CaCl_2$ and $Na_2CO_3$ at a concentration of $0.33M$, at the same time
a vial of $0.33M\ MgSO_4$ was prepared and set aside for later. First
a small vial was filled with $1.5mL$ of $CaCl_2$ followed by $60\mu L$ 
and $90\mu L$ of $MgSO_4$ and $NaCO_3$ respectively, forming a seed solution. 
Next, a larger vial was filled with $5\ mL$, $1.5\ mL$, and $1\ mL$ of
$CaCL_2$, $MgSO_4$, and $NaCO_3$ respectively followed by the seed solution. 
After 10 minutes of slow but continuous mixing a few drops of Agepon was added
to halt the reaction, the solution was filtered and washed 3 times with 
distilled water before being suspended in water. When trapped in circularly 
polarised light, the anisotropic scattering of the sphere results in a periodic 
signal on the QPD. Therefore, the resulting power spectrum is not a Lorentzian
but now also displays peaks that appear at integer multiples of the particles
rotational frequency.

\begin{figure}[h!]
	\label{fig:vaterite}
	\centering
	\begin{subfigure}{0.4\linewidth}
		\includegraphics[width=\linewidth]{vaterite_sample.jpg}
		\subcaption{}
	\end{subfigure}
	\begin{subfigure}{0.55\linewidth}
		\includegraphics[width=\linewidth]{rotating_psd.png}
		\subcaption{}
	\end{subfigure}
	\caption{(a)Sample vaterite sphere suspended in water and trapped by circular polarised trap. (b) Collected power spectrum from rotating vaterite, peaks in the power spectrum appear at integer multiples of the rotational frequency ($f_{rot} \approx 49.8$ Hz)}
\end{figure}

Liquid crystals are an intriguing example of materials with mixed phase
properties, unlike typical solutes such as Glycine, a liquid crystal
can still maintain some degree of order between its individual molecules
while in the liquid state. This is due to the fact that liquid crystals
are constructed of long ordered molecules that demonstrate a long range
ordering. There are three main types of liquid crystal transition methods:
thermotropic crystals will transition to their liquid crystal phase when 
sufficiently heated; Lyotropic materials can undergo this transition due 
to changes in temperature and concentration; and lastly Metallotropic 
materials - which are composed of both organic and inorganic molecules - 
change phase according to the ratio of organic to inorganic molecules present.
Liquid crystal rotors are rather simple in their production, 
4-Heptyl-4-biphenylcarbonitrile (7CB) was purchased from Sigma Aldrich
and a small amount was added to a vial of distilled water. The solution 
was then heated in a water bath to $25^\circ$ in order to transition the
solid crystal into its liquid crystal state. The individual droplets 
are inherently birefringent and rotate in a circularly polarised trap. 
\begin{figure}[h!]
	\centering
	\includegraphics[width=0.5\linewidth]{LC_sample.png}
	\caption{Liquid crystal undergoing rotation due to the circularly polarised trap.} 
\end{figure}

The liquid crystal droplets had a much faster rotation rate than comparable 
vaterite spheres, due to their higher degree of birefringence and the fact
that the droplets are far closer to perfect spheres making angular momentum
transfer more efficient.

\subsection{Rotation of birefringent micro spheres} 
Optical tweezing has often been used for micro-rheology, by computing the
exact forces being exerted on the trapped sphere, one can determine the
local temperature/viscosity of the medium \cite{Millen2014, RodriguezSevilla2018}.
Using a birefringent particle and rotating it within the fluid, the maximum 
rotation rate is due to the fluid drag resisting the torque of the 
trapping beam \cite{RodriguezSevilla2018}. If you want to measure 
fluid flow you can instead use a micro-rotor to see how fluid flow 
propagates in the medium \cite{Knoener2005}. Likewise, one can use 
a galvanometric mirror to probe the drag force of the fluid, by 
understanding the trap strength (calibrating using a low frequency 
signal) one can measure the drag force experienced by the local fluid \cite{RobertsonAnderson2018}. I

@@

From Fig.\ref{fig:vaterite_shear} there is not a strong relationship between
particle size and rotation rate, this is contrary to much of the theoretical
predictions that predict an exponential decay with particle size. This can
be in part due to the fact that synthesising perfectly spherical spheres that
have uniform birefringence is difficult over multiple experiments and tests. 
Despite our best efforts at controlling the growth rate the vaterite spheres
would often combine together while suspended in water after a short period of 
time. The fastest reported rotation rate found during this PhD was by \cite{Arita2016}
that achieved a rotation rate of $5\ kHz$, this is plotted on Fig~\ref{fig:vaterite_shear}
as the dotted line. Even at that extreme a rotation rate the region in which 
nucleation is at its optimal likelihood is only $20\ nm$ wide meaning that 
localising nucleation around a rotating sphere is vanishingly small.

\section{Rotation of micro-rotors in supersaturated solutions}
Vaterite samples were synthesised according to \cite{Parkin2009, Bishop2004} 
(see sec.~\ref{sec:vaterite}), and then suspended in a droplet of 
supersaturated solutions of Glycine and water. After locating a 
singular microsphere the vaterite was trapped in a circularly 
polarised light and brought close to the droplet edge, after a 
period of ten minutes if no nucleation event was observed the 
particle was released and another particle was trapped. Due to 
the increased viscosity of the supersaturated solution and the 
proximity to the droplet edge, this was done as previous work with
optical tweezers have only seen nucleation near the solution-air interface
\cite{Gowayed2021, Liao2022, Walton2018}. The observed rotation rate 
was rather low or non-existent, the results are tabulated below:
\begin{table}[h!]
	\centering
	\caption{Results from rotating vaterite within supersaturated solution of $H_2O$ and Glycine. Solubility concentration for Glycine at $16^\circ$ was $C^*=0.2016g/g$}
	\begin{tabular}[width=\textwidth]{|c|c|c|c|}
		\hline
		Super Saturation & Particle radius [$\mu$] & $\omega$ [Hz] & Nucleation [$\checkmark/\times$]\\
		\hline
		\multirow{3}*{1.01} & 2.34 & 20.7 & $\times$ \\
		\cline{2-4} & 5.67 & 23.3 & $\times$ \\
		\cline{2-4} & 3.26 & 15.4 & $\times$ \\
		\hline
		\multirow{3}*{1.14} & 1.89 & 1.23 & $\times$ \\
		\cline{2-4} & 3.75 & 3.54 & $\times$ \\
		\cline{2-4} & 4.35 & 4.86 & $\times$ \\
		\hline
		\multirow{3}*{1.29} & 3.47 & 0.00 & $\times$ \\
		\cline{2-4} & 1.59 & 0.00 & $\times$ \\
		\cline{2-4} & 6.24 & 0.00 & $\times$ \\
		\hline
		\multirow{3}*{1.40} & 6.32 & 0.00 & $\times$ \\
		\cline{2-4} & 3.68 & 0.00 & $\times$ \\
		\cline{2-4} & 5.43 & 0.00 & $\times$ \\
		\hline
		\multirow{3}*{1.49} & 4.76 & $0.00$ & $\times$ \\
		\cline{2-4} & 7.27 & $0.00$ & $\times$ \\
		\cline{2-4} & 1.52 & $0.00$ & $\times$ \\
		\hline
	\end{tabular}
\end{table}

The calculations from fig.\ref{fig:vaterite_shear} indicate that even
in pure water the rotational speed achieved by a microsphere is
insufficient to achieve a significant shear rate more than a few tenths of a microns
from its surface. The closest we could trap a microsphere to the droplet
edge was in the range of $5-10\mu\ m$, at that distance the fluid flow
would be so low that not even using a liquid droplet rotor would achieve
the rotational speeds necessary to localise nucleation. While in theory
a sufficiently focused laser could rotate any microsphere to a fast enough 
to match the speeds expected by \cite{Debuysschere2023} the localised intensity
would be so large that even using $D_2O$ would see a significant increase in 
temperature.

\section{Shearing via a Galvano-mirror}
Rather than utilising birefringence rotation, which is subject to issues
during synthesis and achieving an ideal particle size, a galvano-mirror
was installed to induce controlled particle motion. While typically 
galvano and gimble mirrors are used to trap multiple particles in a 
regular pattern, in a sufficiently dilute solution a single silica 
micro sphere can be moved quickly through a fluid along a preset path.
Calculating the shear rate around an individual particle is difficult to
do precisely but for low Reynolds numbers we can get an adequate approximation.

For a simple circular path one can estimate the sphere's speed by the 
radius of its path and the frequency of its orbit $U = R\omega$; however 
for a more complex path, such as an elliptical orbit the curve needs 
to be parametrised. One can describe the position parameter of an ellipse 
as such:

@@

where a and b are the different characteristic radii of an ellipse, if 
we say that $u$ describes time from some initial point we can say $u=t\omega$ 
where omega is the frequency of orbit. Substituting this in and then taking 
the partial derivative of position gives:

@@

In order to compute U we simply take the magnitude of our velocity. 
For low velocities the fluid flow at the sphere's surface can be computed 
based on its velocity.

@@

\section{Nucleation using a moving beam}
As mentioned previously, shearing via optical rotation and particle displacement
did not result in
